On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2

### 2023-12-06 manuscript/pnas_revision.tex
changes in response to quality control check at pnas
final manuscript file with appendix and references to appendix removed

\input{make-edits}

@@

% Please include corresponding author, author contribution and author declaration information
% \authorcontributions{Please provide details of author contributions here.}
% \authordeclaration{Please declare any competing interests here.}
% \equalauthors{\textsuperscript{1}A.O.(Author One) contributed equally to this work with A.T. (Author Two) (remove if not applicable).}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: aholbroo@g.ucla.edu}

@@

Within the phylogenetic CTMC models of Section \ref{sec:sars:cov:app}, log-likelihood derivative computations that require $\nabla_{\JJ} e^{t \QQ}$ balloon to $\order{KNd^5}$, for $N$ the number of biological specimens observed and $K$ the number of parameters parameterizing $\QQ$.
To address this overwhelming computational cost, \cite{magee2023random} leverage
the simplistic approximation obtained by setting $n=0$ within \eqref{eq:grad:series}:
\begin{align}\label{eq:firstOrder}
	\widetilde{\nabla}_{\JJ} e^{t\QQ} := t e^{t\QQ}\JJ   \, .
\end{align}
\cite{magee2023random} show that this approximation helps reduce total cost to $\order{Kd^2 + Nd^3}$ and use this speedup within
surrogate-trajectory Hamiltonian Monte Carlo (see \cref{sec:Sur:trad:HMC}) 
to obtain a
34-fold improvement in effective sample size per second (ESS/s) over
random-walk MCMC within their 14 region phylogeographic example.
When trying to explain the remarkable empirical performance of the naive
approximation, the authors derive an error upper
bound (for an arbitrary matrix norm)
\begin{align}\label{eq:initial:bnd}
  \lVert \widetilde{\nabla}_{\JJ}e^{t \QQ}  - \nabla_{\JJ} e^{t \QQ} \rVert
  \leq \frac{ \| \JJ \| \lVert\QQ \rVert }{2} (e^{2t} -2t-1) 
\end{align}
that fails to leverage the specific forms of $\QQ$ and $\JJ$.
Notably, this bound explodes as either $t$ or $\lVert \QQ \rVert$
diverges to $\infty$.  Of course, the latter quantity would be
expected to grow large with dimension $d$ without more careful
structural assumptions, e.g., that $\QQ$ is a rate matrix belonging to
the class
\begin{align}\label{eq:rate:mat}
  \mathcal{R}(d) :=
  \Big\{ \QQ \in {\mathcal M}(d) |
  \QQ_{ij} \geq 0 \text{ for } j \not= i ,\, \sum_{j =1}^d \QQ_{ij} = 0 \Big\} \, .
\end{align}

@@

 \cref{sec:emp}, contains simulation studies comparing: accuracy of matrix exponential derivative approximations for different distributional assumptions on the generator matrix; the posterior distributions obtained using surrogate-trajectory Hamiltonian Monte Carlo \cref{sec:Sur:trad:HMC} and traditional Hamiltonian Monte Carlo; and parameter identification under different priors on generator matrix elements.

@@

This section lays out our rigorous results \cref{thm:ass:error},
\cref{thm:bride:of:frankinstein},
\cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.} and
\cref{cor:rig:det:rand:wrapup}, the proofs of which appear in the
supplemental material below.

@@

These observations lead to the following theorem, the proof of which appears
in the supplement as \cref{sec:Thm:det:main}.
\begin{theorem}\label{thm:ass:error}
  Suppose that $\QQ, \JJ \in \mathcal{M}(d)$ for some $d \geq 1$.  We
  assume that we can find an element $\QQ^+ \in \mathcal{M}(d)$ such that
  $\QQ$ is a generalized inverse of $\QQ^+$, namely,
  \begin{align}\label{eq:GI:prop}
    \QQ^+ \QQ \QQ^+ = \QQ^+
  \end{align}
  and such that
  \begin{align}\label{eq:MP:proj:prop}
    e^{\tau \QQ}(\II - \QQ^+ \QQ) \!=\!  \II - \QQ^+ \QQ, \;
    (\II - \QQ \QQ^+)e^{\tau \QQ} \!=\!  \II - \QQ \QQ^+,
  \end{align}
  for any $\tau \in \bbR$. Furthermore, we suppose that $\QQ^+$ and
  $\QQ$ commute
  \begin{align}\label{eq:GI:com}
    \QQ \QQ^+ = \QQ^+ \QQ.
  \end{align}
  Finally, taking $\|\cdot\|$ be any matrix norm, we assume that
  \begin{align}\label{eq:qetq:exp:decay}
    \| \QQ e^{\tau \QQ}\| \leq C_0 e^{-\kappa \tau},
    \quad
    \text{ for all } \tau \geq 0,
  \end{align}
  where the constants $C_0 > 0, \kappa > 0$
  are independent of $\tau$.  Then, under these circumstances, 
  \begin{align}\label{eq:ass:error}
    \|\QQ^+ \JJ (\II - \QQ \QQ^{+})-& (\II -\QQ^+ \QQ) \JJ \QQ^{+}
        +t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
                \notag\\
      &
        +  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ\| \leq C(1+ t)e^{-\kappa t}
\end{align}
for any $t \geq 0$.  Here, $C > 0$ is a $t$-independent
constant which is given explicitly as
\begin{align}
  C_0 \big(\| (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\|
  \!&+\! \|(\QQ^+)^2\JJ (\II - \QQ \QQ^{+})\|
    \!+\! \|\QQ^+ \hspace{0.5mm}\JJ\|\big)
    \notag\\
    &+ C_0^2 \| \QQ^+  \JJ  \QQ^{+} \|.
      \label{eq:exp:error:ass:er}
  \end{align}
\end{theorem}

@@

We formulate our second major result as follows. 
\begin{theorem}\label{t:lambda_n-1(Q)->-infty_non-symm_a.s.}
  Let $F_X \in L_{\psi_1}$ be a distribution such that $X \geq 0$
  a.s., $\bbE X = \mu > 0$ and $\Var X = \sigma^2 > 0$. We consider a
  sequence of random matrices
  $\QQ \equiv \QQ(d) = \{q_{ij}\}_{i,j=1,\hdots,d} \in \mathcal{R}(d)$, $d \in \bbN$,
  where either
  \begin{align}\label{e:RM_theorem_general_condition}
    \begin{aligned}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i \neq j}
    \stackrel{\textnormal{iid}}\sim F_X
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d
  \end{aligned}
  \end{align}
  or we impose that $\QQ \in \mathcal{R}(d) \cap \mathcal{S}(d)$ as 
    \begin{align}
    \begin{aligned}\label{e:RM_theorem_symmetry_condition}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i > j}
  \stackrel{\textnormal{iid}}\sim F_X,
  q_{ij} := q_{ji} \text{ for } i < j
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d.
  \end{aligned}
\end{align}
    Then, in either of these cases, for any $d \in \bbN \backslash\{1\}$, we have
    \begin{align}
 \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big) \leq& \frac{\sigma_2(\QQ)}{d} 
 \notag \\
 \leq& \frac{\sigma_d(\QQ)}{d}
 \leq \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)
 \label{e:lambda_n-1(Q)->-infty_non-symm}
      \end{align}
      almost surely. In \eqref{e:lambda_n-1(Q)->-infty_non-symm},
      $O_{\textnormal{a.s.}}$ is as in
      \eqref{e:X=O_a.s.(1)_Y=o_a.s.(1)}.
\end{theorem}

@@

\begin{figure}[!t]
	\centering
	\includegraphics[width=\linewidth]{lowDimVis.pdf}
	\vspace{-2em}
	\caption{Posterior density plots for surrogate-trajectory Hamiltonian Monte Carlo (HMC) (\cref{sec:Sur:trad:HMC}) using the naive approximate derivative, the corrected approximation \eqref{eq:better:help:on:sale} and the exact matrix exponential derivative for elements of a $5\times 5$ generator matrix. The near perfect overlap reflects the fact that each algorithm's transition kernel leaves the posterior distribution invariant \cite{glatt2020accept}.  To generate data, we randomly draw standard normal generator entries once and simulate 20 independent initial/final position pairs from a CTMC with time interval $t=1$.  We show the true value in red and negate diagonal elements to simplify presentation.  }\label{fig:lowDim}
\end{figure}

@@

\cite{magee2023random} consider phylogenetic models that involve CTMC
priors and show that the first-order approximation
\eqref{eq:firstOrder} of the matrix exponential derivative
\eqref{eq:grad:def} performs remarkably well within surrogate HMC (see
\cref{sec:Sur:trad:HMC} in the Supplement), achieving an over 30-fold
efficiency gain compared to random-walk Metropolis for a 14-state
model with over 180 model parameters.  Here, we demonstrate similar
strong performance of the first-order approximation within surrogate
HMC for a 44-state model with almost 1,900 model parameters.

@@

\appendix

\section{Proof of \cref{thm:ass:error}}
\label{sec:Thm:det:main}

Recall that $\Z(t)$ is defined as in \eqref{eq:true:diff}, and
consider the representation \eqref{eq:ass:procs:0}. We begin by
decomposing the first term $e^{-s\QQ} \JJ e^{s \QQ}$ inside of the
integral in \eqref{eq:ass:procs:0} as
\begin{align}
  e^{-s\QQ} &\JJ e^{s \QQ} \notag\\
  =&e^{-s\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}
     + e^{-s\QQ} (\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
                  \notag\\
      &+ e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+}) e^{s\QQ}
                \notag\\
       &+e^{-s\QQ} (\II -\QQ^+ \QQ) \JJ (\II -\QQ \QQ^{+}) e^{s\QQ}
                           \notag\\
  = &e^{-s\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}
      + (\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
         \label{eq:big:yuck:1}                          
      \\
      &+ e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})
      %\notag\\
  + (\II -\QQ^+ \QQ) \JJ (\II -\QQ \QQ^{+}),
  \notag
\end{align}
where we used \eqref{eq:MP:proj:prop} for the second identity.
Regarding the middle two terms, with our commutativity
\eqref{eq:GI:com} assumption, we observe that
\begin{align}
  (\II &-\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
         + e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})
                \label{eq:big:yuck:2}                          
                     \\
  &=  (\II -\QQ^+ \QQ) \JJ \QQ^{+} \QQ e^{s\QQ}
    + e^{-s\QQ}\QQ \QQ^+ \JJ (\II - \QQ \QQ^{+})
    \notag\\
  &= \frac{d}{ds}
    \left(
    (\II -\QQ^+ \QQ) \JJ \QQ^{+} e^{s\QQ}
    - e^{-s\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
    \right).
 \notag
\end{align}
Hence, integrating in time, and then using \eqref{eq:GI:prop},
\eqref{eq:GI:com} for the first and last term terms on the right hand side, we have
\begin{align}
  \int_0^t& \big\{(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
                             + e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})\big\}ds
                             \notag\\
   =& (\II -\QQ^+ \QQ) \JJ \QQ^{+}e^{t\QQ}
      - e^{-t\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
      \notag\\
    &- (\II -\QQ^+ \QQ) \JJ \QQ^{+} 
      +  \QQ^+ \JJ (\II - \QQ \QQ^{+})
      \notag\\
   =& (\II -\QQ^+ \QQ) \JJ (\QQ^+)^2 \QQ e^{t\QQ}
      - e^{-t\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
      \notag\\
    &- (\II -\QQ^+ \QQ) \JJ \QQ^{+} 
      + \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+})
             \label{eq:big:yuck:3}                          
\end{align}
Finally let us note, regarding the second term on the right hand side
of \eqref{eq:ass:procs:0}, using again 
\eqref{eq:MP:proj:prop} and \eqref{eq:GI:prop}, \eqref{eq:GI:com}
\begin{align}
  -te^{t\QQ}\JJ 
  =-te^{t\QQ}\QQ \QQ^+ \JJ  - t (\II - \QQ^+\QQ) \JJ.
               \label{eq:small:yuck:4}                          
\end{align}

By combining \eqref{eq:big:yuck:1}--\eqref{eq:small:yuck:4} and
comparing with \eqref{eq:ass:procs:0}, we now find that
\begin{align}\label{e:E(t)=integ+sum}
  \Z(t)\! =\! \! \int_0^t \! \! \! e^{(t-s)\QQ} &\QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}ds
           + (\II - \!\QQ^+ \QQ) \JJ (\QQ^{+})^2\QQ e^{t\QQ}
           \notag\\
           &+ e^{t\QQ} \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+}) -te^{t\QQ}\QQ \QQ^+ \JJ 
           \notag\\
           &-\QQ^+ \JJ (\II - \QQ \QQ^{+})- (\II -\QQ^+ \QQ) \JJ \QQ^{+}
             \notag\\
         &-t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}.
\end{align}
Rearranging \eqref{e:E(t)=integ+sum}, we have 
\begin{align*}
\QQ^+ \JJ (\II - \QQ \QQ^{+}) &+(\II -\QQ^+ \QQ) \JJ \QQ^{+}
   +t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
                                \notag\\
  &+  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ
  = \TT_1 + \TT_2,
\end{align*}
where 
\begin{align*}
 \TT_1 :=&  \int_0^t  \! \! e^{(t-s)\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}ds,\\
  \TT_2 :=&  (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\QQ e^{t\QQ}\\
         &+ e^{t\QQ} \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+}) -te^{t\QQ}\QQ \QQ^+ \JJ.
\end{align*}

We estimate each of $\TT_1$ and $\TT_2$ in turn.  Regarding $\TT_1$,
using once more \eqref{eq:GI:com}, as well as \eqref{eq:qetq:exp:decay}, we have
\begin{align*}
  \| \TT_1\| \leq&
    \int_0^t \|\QQ e^{(t-s)\QQ}\| \| \QQ^+  \JJ \QQ^{+} \|\| \QQ e^{s\QQ}\|ds
         \notag\\
   \leq& C_0^2\int_0^t  e^{-(t-s)\kappa} \| \QQ^+  \JJ \QQ^{+} \| e^{-s\kappa } ds
                         \notag\\
   =& C_0^2 t e^{-t\kappa} \| \QQ^+  \JJ  \QQ^{+} \|.
\end{align*}
Turning to $\TT_2$, by \eqref{eq:qetq:exp:decay} we have
\begin{align*}
  \| \TT_2\|
  \leq C_0 (&\| (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\|
               + \|(\QQ^+)^2\JJ (\II - \QQ \QQ^{+})\|
               \notag\\
               &+ t\|\QQ^+ \hspace{0.25mm}\JJ\|) e^{- t \kappa }.
\end{align*}
Combining these two bound completes the proof.

\section{Proof of \cref{thm:bride:of:frankinstein}}

>>>>

%\input{make-edits}

@@

% Please include corresponding author, author contribution and author declaration information
 \authorcontributions{AJH and MAS designed research. GD, NEGH and AJH developed mathematical results. AJH and AFM performed simulations and data analysis.  GD, NEGH and AJH wrote the paper.}
% \authordeclaration{Please declare any competing interests here.}
% \equalauthors{\textsuperscript{1}A.O.(Author One) contributed equally to this work with A.T. (Author Two) (remove if not applicable).}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: aholbroo@g.ucla.edu}

@@

Within the phylogenetic CTMC models of Section \ref{sec:sars:cov:app}, log-likelihood derivative computations that require $\nabla_{\JJ} e^{t \QQ}$ balloon to $\order{KNd^5}$, for $N$ the number of biological specimens observed and $K$ the number of parameters parameterizing $\QQ$.
To address this overwhelming computational cost, \cite{magee2023random} leverage
the simplistic approximation obtained by setting $n=0$ within \eqref{eq:grad:series}:
\begin{align}\label{eq:firstOrder}
	\widetilde{\nabla}_{\JJ} e^{t\QQ} := t e^{t\QQ}\JJ   \, .
\end{align}
\cite{magee2023random} show that this approximation helps reduce total cost to $\order{Kd^2 + Nd^3}$ and use this speedup within
surrogate-trajectory Hamiltonian Monte Carlo (see Supplement) 
to obtain a
34-fold improvement in effective sample size per second (ESS/s) over
random-walk MCMC within their 14 region phylogeographic example.
When trying to explain the remarkable empirical performance of the naive
approximation, the authors derive an error upper
bound (for an arbitrary matrix norm)
\begin{align}\label{eq:initial:bnd}
  \lVert \widetilde{\nabla}_{\JJ}e^{t \QQ}  - \nabla_{\JJ} e^{t \QQ} \rVert
  \leq \frac{ \| \JJ \| \lVert\QQ \rVert }{2} (e^{2t} -2t-1) 
\end{align}
that fails to leverage the specific forms of $\QQ$ and $\JJ$.
Notably, this bound explodes as either $t$ or $\lVert \QQ \rVert$
diverges to $\infty$.  Of course, the latter quantity would be
expected to grow large with dimension $d$ without more careful
structural assumptions, e.g., that $\QQ$ is a rate matrix belonging to
the class
\begin{align}\label{eq:rate:mat}
  \mathcal{R}(d) :=
  \Big\{ \QQ \in {\mathcal M}(d) |
  \QQ_{ij} \geq 0 \text{ for } j \not= i ,\, \sum_{j =1}^d \QQ_{ij} = 0 \Big\} \, .
\end{align}

@@

 \cref{sec:emp} contains simulation studies comparing: accuracy of matrix exponential derivative approximations for different distributional assumptions on the generator matrix; the posterior distributions obtained using surrogate-trajectory Hamiltonian Monte Carlo and traditional Hamiltonian Monte Carlo; and parameter identification under different priors on generator matrix elements.

@@

This section lays out our rigorous results \cref{thm:ass:error},
\cref{thm:bride:of:frankinstein},
\cref{t:lambda_n-1(Q)->-infty_non-symm_a.s.} and
\cref{cor:rig:det:rand:wrapup}, the proofs of which appear in the
Supplement.

@@

These observations lead to the following theorem, the proof of which appears
in the Supplement.
\begin{theorem}\label{thm:ass:error}
  Suppose that $\QQ, \JJ \in \mathcal{M}(d)$ for some $d \geq 1$.  We
  assume that we can find an element $\QQ^+ \in \mathcal{M}(d)$ such that
  $\QQ$ is a generalized inverse of $\QQ^+$, namely,
  \begin{align}\label{eq:GI:prop}
    \QQ^+ \QQ \QQ^+ = \QQ^+
  \end{align}
  and such that
  \begin{align}\label{eq:MP:proj:prop}
    e^{\tau \QQ}(\II - \QQ^+ \QQ) \!=\!  \II - \QQ^+ \QQ, \;
    (\II - \QQ \QQ^+)e^{\tau \QQ} \!=\!  \II - \QQ \QQ^+,
  \end{align}
  for any $\tau \in \bbR$. Furthermore, we suppose that $\QQ^+$ and
  $\QQ$ commute
  \begin{align}\label{eq:GI:com}
    \QQ \QQ^+ = \QQ^+ \QQ.
  \end{align}
  Finally, taking $\|\cdot\|$ be any matrix norm, we assume that
  \begin{align}\label{eq:qetq:exp:decay}
    \| \QQ e^{\tau \QQ}\| \leq C_0 e^{-\kappa \tau},
    \quad
    \text{ for all } \tau \geq 0,
  \end{align}
  where the constants $C_0 > 0, \kappa > 0$
  are independent of $\tau$.  Then, under these circumstances, 
  \begin{align}\label{eq:ass:error}
    \|\QQ^+ \JJ (\II - \QQ \QQ^{+})-& (\II -\QQ^+ \QQ) \JJ \QQ^{+}
        +t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
                \notag\\
      &
        +  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ\| \leq C(1+ t)e^{-\kappa t}
\end{align}
for any $t \geq 0$.  Here, $C > 0$ is a $t$-independent
constant which is given explicitly as
\begin{align}
  C_0 \big(\| (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\|
  \!&+\! \|(\QQ^+)^2\JJ (\II - \QQ \QQ^{+})\|
    \!+\! \|\QQ^+ \hspace{0.5mm}\JJ\|\big)
    \notag\\
    &+ C_0^2 \| \QQ^+  \JJ  \QQ^{+} \|.
      \label{eq:exp:error:ass:er}
  \end{align}
\end{theorem}

@@

We formulate our second major result as follows. 
\begin{theorem}\label{t:lambda_n-1(Q)->-infty_non-symm_a.s.}
  Let $F_X \in L_{\psi_1}$ be a distribution such that $X \geq 0$
  a.s., $\bbE X = \mu > 0$ and $\Var X = \sigma^2 > 0$. We consider a
  sequence of random matrices
  $\QQ \equiv \QQ(d) = \{q_{ij}\}_{i,j=1,\hdots,d} \in \mathcal{R}(d)$, $d \in \bbN$,
  where either
  \begin{align}\label{e:RM_theorem_general_condition}
    \begin{aligned}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i \neq j}
    \stackrel{\textnormal{iid}}\sim F_X
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d
  \end{aligned}
  \end{align}
  or we impose that $\QQ \in \mathcal{R}(d) \cap \mathcal{S}(d)$ as 
    \begin{align}
    \begin{aligned}\label{e:RM_theorem_symmetry_condition}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i > j}
  \stackrel{\textnormal{iid}}\sim F_X,
  q_{ij} := q_{ji} \text{ for } i < j
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d.
  \end{aligned}
\end{align}
    Then, in either of these cases, for any $d \in \bbN \backslash\{1\}$, we have
    \begin{align}
 \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big) \leq& \frac{\sigma_2(\QQ)}{d} 
 \notag \\
 \leq& \frac{\sigma_d(\QQ)}{d}
 \leq \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)
 \label{e:lambda_n-1(Q)->-infty_non-symm}
      \end{align}
      almost surely. %In \eqref{e:lambda_n-1(Q)->-infty_non-symm},
      %$O_{\textnormal{a.s.}}$ is as in
      %\eqref{e:X=O_a.s.(1)_Y=o_a.s.(1)}.
\end{theorem}

@@

\begin{figure}[!t]
	\centering
	\includegraphics[width=\linewidth]{lowDimVis.pdf}
	\vspace{-2em}
	\caption{Posterior density plots for surrogate-trajectory Hamiltonian Monte Carlo (HMC) using the naive approximate derivative, the corrected approximation \eqref{eq:better:help:on:sale} and the exact matrix exponential derivative for elements of a $5\times 5$ generator matrix. The near perfect overlap reflects the fact that each algorithm's transition kernel leaves the posterior distribution invariant \cite{glatt2020accept}.  To generate data, we randomly draw standard normal generator entries once and simulate 20 independent initial/final position pairs from a CTMC with time interval $t=1$.  We show the true value in red and negate diagonal elements to simplify presentation.  }\label{fig:lowDim}
\end{figure}

@@

\cite{magee2023random} consider phylogenetic models that involve CTMC
priors and show that the first-order approximation
\eqref{eq:firstOrder} of the matrix exponential derivative
\eqref{eq:grad:def} performs remarkably well within surrogate HMC (see Supplement), achieving an over 30-fold
efficiency gain compared to random-walk Metropolis for a 14-state
model with over 180 model parameters.  Here, we demonstrate similar
strong performance of the first-order approximation within surrogate
HMC for a 44-state model with almost 1,900 model parameters.

@@

%\appendix  
%
%
%\section{Proof of \cref{thm:ass:error}}
%\label{sec:Thm:det:main}
%
%
%Recall that $\Z(t)$ is defined as in \eqref{eq:true:diff}, and
%consider the representation \eqref{eq:ass:procs:0}. We begin by
%decomposing the first term $e^{-s\QQ} \JJ e^{s \QQ}$ inside of the
%integral in \eqref{eq:ass:procs:0} as
%\begin{align}
%  e^{-s\QQ} &\JJ e^{s \QQ} \notag\\
%  =&e^{-s\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}
%     + e^{-s\QQ} (\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
%                  \notag\\
%      &+ e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+}) e^{s\QQ}
%                \notag\\
%       &+e^{-s\QQ} (\II -\QQ^+ \QQ) \JJ (\II -\QQ \QQ^{+}) e^{s\QQ}
%                           \notag\\
%  = &e^{-s\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}
%      + (\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
%         \label{eq:big:yuck:1}                          
%      \\
%      &+ e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})
%      %\notag\\
%  + (\II -\QQ^+ \QQ) \JJ (\II -\QQ \QQ^{+}),
%  \notag
%\end{align}
%where we used \eqref{eq:MP:proj:prop} for the second identity.
%Regarding the middle two terms, with our commutativity
%\eqref{eq:GI:com} assumption, we observe that
%\begin{align}
%  (\II &-\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
%         + e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})
%                \label{eq:big:yuck:2}                          
%                     \\
%  &=  (\II -\QQ^+ \QQ) \JJ \QQ^{+} \QQ e^{s\QQ}
%    + e^{-s\QQ}\QQ \QQ^+ \JJ (\II - \QQ \QQ^{+})
%    \notag\\
%  &= \frac{d}{ds}
%    \left(
%    (\II -\QQ^+ \QQ) \JJ \QQ^{+} e^{s\QQ}
%    - e^{-s\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
%    \right).
% \notag
%\end{align}
%Hence, integrating in time, and then using \eqref{eq:GI:prop},
%\eqref{eq:GI:com} for the first and last term terms on the right hand side, we have
%\begin{align}
%  \int_0^t& \big\{(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+} e^{s\QQ}
%                             + e^{-s\QQ}\QQ^+ \QQ \JJ (\II - \QQ \QQ^{+})\big\}ds
%                             \notag\\
%   =& (\II -\QQ^+ \QQ) \JJ \QQ^{+}e^{t\QQ}
%      - e^{-t\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
%      \notag\\
%    &- (\II -\QQ^+ \QQ) \JJ \QQ^{+} 
%      +  \QQ^+ \JJ (\II - \QQ \QQ^{+})
%      \notag\\
%   =& (\II -\QQ^+ \QQ) \JJ (\QQ^+)^2 \QQ e^{t\QQ}
%      - e^{-t\QQ} \QQ^+ \JJ (\II - \QQ \QQ^{+})
%      \notag\\
%    &- (\II -\QQ^+ \QQ) \JJ \QQ^{+} 
%      + \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+})
%             \label{eq:big:yuck:3}                          
%\end{align}
%Finally let us note, regarding the second term on the right hand side
%of \eqref{eq:ass:procs:0}, using again 
%\eqref{eq:MP:proj:prop} and \eqref{eq:GI:prop}, \eqref{eq:GI:com}
%\begin{align}
%  -te^{t\QQ}\JJ 
%  =-te^{t\QQ}\QQ \QQ^+ \JJ  - t (\II - \QQ^+\QQ) \JJ.
%               \label{eq:small:yuck:4}                          
%\end{align}
%
%By combining \eqref{eq:big:yuck:1}--\eqref{eq:small:yuck:4} and
%comparing with \eqref{eq:ass:procs:0}, we now find that
%\begin{align}\label{e:E(t)=integ+sum}
%  \Z(t)\! =\! \! \int_0^t \! \! \! e^{(t-s)\QQ} &\QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}ds
%           + (\II - \!\QQ^+ \QQ) \JJ (\QQ^{+})^2\QQ e^{t\QQ}
%           \notag\\
%           &+ e^{t\QQ} \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+}) -te^{t\QQ}\QQ \QQ^+ \JJ 
%           \notag\\
%           &-\QQ^+ \JJ (\II - \QQ \QQ^{+})- (\II -\QQ^+ \QQ) \JJ \QQ^{+}
%             \notag\\
%         &-t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}.
%\end{align}
%Rearranging \eqref{e:E(t)=integ+sum}, we have 
%\begin{align*}
%\QQ^+ \JJ (\II - \QQ \QQ^{+}) &+(\II -\QQ^+ \QQ) \JJ \QQ^{+}
%   +t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}
%                                \notag\\
%  &+  \nabla_\JJ e^{t \QQ} - t  e^{t\QQ} \JJ
%  = \TT_1 + \TT_2,
%\end{align*}
%where 
%\begin{align*}
% \TT_1 :=&  \int_0^t  \! \! e^{(t-s)\QQ} \QQ^+ \QQ \JJ \QQ \QQ^{+} e^{s\QQ}ds,\\
%  \TT_2 :=&  (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\QQ e^{t\QQ}\\
%         &+ e^{t\QQ} \QQ (\QQ^+)^2 \JJ (\II - \QQ \QQ^{+}) -te^{t\QQ}\QQ \QQ^+ \JJ.
%\end{align*}
%
%We estimate each of $\TT_1$ and $\TT_2$ in turn.  Regarding $\TT_1$,
%using once more \eqref{eq:GI:com}, as well as \eqref{eq:qetq:exp:decay}, we have
%\begin{align*}
%  \| \TT_1\| \leq&
%    \int_0^t \|\QQ e^{(t-s)\QQ}\| \| \QQ^+  \JJ \QQ^{+} \|\| \QQ e^{s\QQ}\|ds
%         \notag\\
%   \leq& C_0^2\int_0^t  e^{-(t-s)\kappa} \| \QQ^+  \JJ \QQ^{+} \| e^{-s\kappa } ds
%                         \notag\\
%   =& C_0^2 t e^{-t\kappa} \| \QQ^+  \JJ  \QQ^{+} \|.
%\end{align*}
%Turning to $\TT_2$, by \eqref{eq:qetq:exp:decay} we have
%\begin{align*}
%  \| \TT_2\|
%  \leq C_0 (&\| (\II -\QQ^+ \QQ) \JJ (\QQ^{+})^2\|
%               + \|(\QQ^+)^2\JJ (\II - \QQ \QQ^{+})\|
%               \notag\\
%               &+ t\|\QQ^+ \hspace{0.25mm}\JJ\|) e^{- t \kappa }.
%\end{align*}
%Combining these two bound completes the proof.
%
%\section{Proof of \cref{thm:bride:of:frankinstein}}
%
%This result is established from \cref{thm:ass:error} and some basic
%properties $\QQ$ under the given assumptions. First, note that, cf.
%\eqref{eq:ev:d:cond:sym:case}, (\ref{eq:EV:real:ass})
%\begin{equation}\label{e:max_|lambda-j(Q)|=<|lambda-1(Q)|}
%  |\lambda_d(\QQ)| \leq \max_{j =1, \ldots,d_{-}} |\lambda_{j}(\QQ)| \leq |\lambda_1(\QQ)|.
%\end{equation}
%Referring back to
%\eqref{eq:frob:norm}, by \eqref{e:max_|lambda-j(Q)|=<|lambda-1(Q)|} we have the estimate
%\begin{align}\label{eq:frob:sucks}
%  \| \QQ e^{t\QQ} \|_F
%  =&  \left( \sum_{j =1}^{d_{-}}
%     | \lambda_j(\QQ) e^{ t  \lambda_j(\QQ)  } |^2 \right)^{1/2}
%     \notag\\
%  \leq& \sqrt{d_{-}} |\lambda_{1}(\QQ)| e^{-t |\lambda_{d_{-}}(\QQ)|}.
%\end{align}
%Similarly, from \eqref{eq:op:norm} and \eqref{e:max_|lambda-j(Q)|=<|lambda-1(Q)|}, we have that
%\begin{align}\label{eq:op:better}
%  \| \QQ e^{t\QQ} \|_{\op}
%  \leq& \max_{j =1, \ldots,d_{-}}
%        |\lambda_{j}(\QQ)| e^{-t |\lambda_{j}(\QQ)|}
%        \notag\\
